Metric Spaces with Expensive Distances

### 2019-08-15 paper/sosa_acm.tex
Add SOSA submission, reformat to ACM

@@

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric \cite{rubner2000earth}.

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see \cref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

@@

\myparagraph{Spanners and known constructions}
%
Let $(\metricspace, \dist)$ be a finite metric space with $n$ points. 
One way to encode the metric space is a complete weighted graph on $\metricspace$,
where the weights correspond to the distances of the points.
A subgraph $G$ of this graph is called a \emph{$(1+\eps)$-spanner} for $(\metricspace,\dist)$ 
if for any pair of points $(u,v)$,
the shortest path distance $d_{uv}$ of $u$ and $v$ in $G$ satisfies $d_{u,v}\leq (1+\eps)\dist(u,v)$.
In other words, the shortest path metric of $G$ is a good approximation of the actual distance for every pair of points.
Clearly, it is a necessary condition that $G$ is connected, hence every spanner must have at least $n-1$ edges.

@@

On the other hand, spanner constructions based on Well-Separated Pair Decomposition\footnote{ 
for convenience, we repeat the definition of WSPD in \cref{sec:appendix_background}}~\cite{cal-kos-wspd,hm-fast} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a \espanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

@@

We performed two types of experiments: on points sampled from a low-dimensional Euclidean
space and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

\cref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners produces
spanners of the same quality as the standard greedy algorithm,
\bgrdy and all variants of \brndm
perform significantly better than both variants of \textsc{BlindQuasiSorted}.

\ignore{
As for different variants of the \brndm algorithm,
we note that their performance is almost the same, and the algorithm works
significantly better than \textsc{QuasiSorted} variants, but obviously
worse than the blind greedy variant. There is a consistent, though small,
difference between the variants that do not force lower bounds first and
the other two variants of the \brndm.  (see \cref{fig:blind_rbr_variants}). 
}

@@

From our experiments, we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners. More data from experiments in $\RR^n$ can be found in
\cref{sec:spanners_euclidean_addendum}.

@@

The other data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape data set; we only used
persitence in dimension 0, since the diagrams in this dimension are larger in size.
For the \dtype{perturbed} data set,
we added random noise to points of the original shapes,
creating 10 modified versions of each of them;
then we computed persistence diagrams in dimension 0 of these modified
shapes.
Persistence diagrams were computed with the \libraryname{DIPHA}
library \cite{bauer2014distributed}. 
Wasserstein distance were computed approximately
with the software library \hera using a relative error of $0.01$.
We restricted to approximate computation because computing the
Wasserstein distance exactly (using the Hungarian algorithm)
has been reported to be very expensice~\cite{kerber2017geometry}.

Sparseness values of \bgrdy spanner on the whole \dtype{original} data set
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
the advantage of our algorithm fades away. A plausible explanation
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in size of the greedy spanner.
For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
computations for $\eps = 0.1$; for a $1.5$-approximation, we
reduce the number of distances computed by a factor of 10.

@@

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
We compared our approach for a fixed
$\eps$ with the time to compute all pairwise-distance between the
persistence diagrams using \hera, using the same approximation parameter
$\eps$. Let us emphasize that the diagrams we have are rather small,
with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time,
and we expect this advantage to grow for diagrams of larger cardinality.

In order to see the growth of the spanner size, we use the \dtype{perturbed}
data set, drawing from it random samples.
We plot the sparseness of the spanners obtained by \grdy
and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
for different values of $\eps$, always using the $\wsdist{1}$ distance.
As expected, the ratio of computed distances decreases as the number of diagrams
grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
the growth of the spanner size relative to the number of diagrams.
For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
linear growth) for larger values of $n$.

@@

The full proof is given in \cref{sec:proof_ann},
we summarize the main ideas:
first, we notice that points ``far away'' from the current candidate $c$ 
can be disregarded in the algorithm. More precisely, if $p_i$
is outside of the ball of radius $2\dist(c, q)$ centered at $c$,
the lower bound for $p_i$ will be so large that
the distance to of $p_i$ to $q$ does not have to be queried.
Second, whenever a distance of $q$ to some $p_j$ is computed,
we can disregard all points ``very close'' to $p_j$.
Precisely, points in a ball of radius $\eps\dist(p_j,q)$
will not invalidate the current candidate 
to be an approximate nearest neighbor and hence,
no distance query for such points is needed.

@@

The proof strategy fails for $\eps=0$ because the $\eps$-ball
around $p_j$ as described above degenerates, and the packing argument fails.
Indeed, as the example in \cref{fig:exact_bad_example} (in~\cref{sec:proof_ann}) 
shows, there are point sets where the expected number
of distance computations for exact nearest neighbour is linear.

@@

>>>>

\documentclass[natbib=false]{acmart}

\usepackage{amsmath}%
\usepackage{amssymb}%
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{booktabs}
\usepackage{nccmath}
\usepackage{diagbox}
\usepackage{url}
\usepackage{xspace}
\usepackage{mathbbol}
\usepackage{comment}
\usepackage{braket}

% graphics
\usepackage{tikz}
\usepackage{pgf}
\usepackage{graphicx}
\usepackage{pgfplots}
\pgfplotsset{compat=1.14}
\usepackage{pgfplotstable}
\usetikzlibrary{patterns}
\usetikzlibrary{external}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{color}
\usepackage{mwe}
\usepackage{floatflt}

\usetikzlibrary{arrows}

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\newcommand{\RR}{\mathbb{R}}

\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\renewcommand{\epsilon}{\varepsilon}
\renewcommand{\phi}{\varphi}
\newcommand{\eps}{\varepsilon}

\newcommand{\ignore}[1]{}
\newcommand{\yaxis}{$y$\nobreakdash-axis\xspace}
\newcommand{\xaxis}{$x$\nobreakdash-axis\xspace}

\newcommand{\algname}[1]{{\textsc{#1}}\xspace}

\newcommand{\dtype}[1]{{\textit{\small #1}}}

\newcommand{\libraryname}[1]{{\textsc{#1}}\xspace}

\newcommand{\hera}{\libraryname{Hera}}
\newcommand{\Hera}{\libraryname{Hera}}
\newcommand{\dionisus}{\libraryname{Dionysus}}
\newcommand{\Dionisus}{\libraryname{Dionysus}}

\pgfplotsset
{
    tick label style={font=\footnotesize},
    legend style    ={font=\footnotesize},
    label style     ={font=\footnotesize},
    tickwidth       =2pt,
    ylabel near ticks,
    xlabel near ticks,
    legend style     ={draw=none},
    legend cell align=right,
    legend plot pos  =right,
    /pgfplots/narrow area legend/.style=
    {
        legend image code/.code={ \draw[#1] (0cm,-0.06cm) rectangle (0.6cm,0.06cm); }
    },
    /pgfplots/short line legend/.style=
    {
        legend image code/.code={ \draw[#1] (0cm,0cm) rectangle (0.6cm,0cm); }
    },
}

\pgfplotscreateplotcyclelist{filled-color}
{
    every mark/.append style={solid,fill=blue!80!white},    mark=*          \\%
    every mark/.append style={solid,fill=red!80!white},     mark=square*    \\%
    every mark/.append style={solid,fill=orange!80!white},  mark=pentagon*  \\%
    every mark/.append style={solid,fill=green},            mark=diamond*   \\%
    every mark/.append style={solid,fill=black},            mark=triangle*  \\%
}

\newcommand{\addaveraged}[2][]
           {\addplot+[#1,error bars/.cd, y dir=both,y explicit] table[x index=0, y index=1, y error index=2] {#2};}

\pgfplotsset{
    geommark/.style={black,mark options={solid,fill=red!80!white},mark=square*},
    nongeommark/.style={black,mark options={solid,fill=blue!80!white},mark=*},
    dionmark/.style={black,mark options={solid,fill=orange!80!white},mark=pentagon*},
    greentrianglemark/.style={black,mark options={solid,fill=green!80!white},mark=triangle*},
    browndiamondmark/.style={black,mark options={solid,fill=brown!80!white},mark=diamond*}
}

% all distances
% interleaving
\newcommand{\intdist}{{D_I}}
% bottleneck
\newcommand{\btdist}{{W_\infty}}
% Wasserstein of order q (parameter)
\newcommand{\wsdist}[1]{{W_{#1}}}

%% \category{dummy}

% \category{dummy}

%\Copyright{Michael Kerber and Arnur Nigmetov}
%\Copyright{Anonymous Authors}
% END uncomment for LIPICS

\def \expDistFigPath {data/expensive_distances/pics/}
\def \expDistDataPath {data/expensive_distances/}
\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}

\renewcommand{\S}{\mathbb{S}}
\newcommand{\dappr}{\tilde{d}}
\newcommand{\orient}{O}
\newcommand{\cpro}{\Pi}

\newcommand{\metricspace}{\mathcal{M}}
\newcommand{\approxmetric}{\gamma}
\newcommand{\parent}{\mathrm{par}}
\newcommand{\point}{p}
\newcommand{\level}{\ell}
\newcommand{\diam}{\mathrm{diam}}
\newcommand{\diag}{{\Delta}}
\newcommand{\pointset}{P}
\newcommand{\distspace}{\mathcal{M}}
\newcommand{\dist}{\delta}
\newcommand{\adist}{\gamma}
\newcommand{\complexity}{C_{\dist}}
\newcommand{\doublingdimension}{\Delta}

\newcommand{\espanner}{{$(1+\eps)$-spanner}}

\newcommand{\grdy}{\algname{Greedy}}
\newcommand{\bgrdy}{\algname{BlindGreedy}}
\newcommand{\brndm}{\algname{BlindRandom}}
\newcommand{\bqsgrdy}{\algname{BlindQuasiSortedGreedy}}
\newcommand{\bqsshaker}{\algname{BlindQuasiSortedShaker}}

\newcommand{\tX}{{\tilde{X}}}
\newcommand{\tY}{{\tilde{Y}}}

\newcommand{\myparagraph}[1]{\textbf{#1.}}

\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}

\def \expDistFigPath {pics/}
\def \expDistDataPath {./}

\usepackage[style=ACM-Reference-Format,backend=bibtex,sorting=none]{biblatex}
\addbibresource{bib.bib}

\title{Metric Spaces with Expensive Distances}

\author{Michael Kerber}
\email{kerber@tugraz.at}
\affiliation{%
  \institution{Graz University of Technology}
  \streetaddress{Kopernikusgasse 24}
  \city{Graz}
  \postcode{8010}
  \country{Austria}
}

\author{Arnur Nigmetov}
\email{nigmetov@tugraz.at}
\affiliation{%
  \institution{Graz University of Technology}
  \streetaddress{Kopernikusgasse 24}
  \city{Graz}
  \postcode{8010}
  \country{Austria}
}

\begin{abstract}
In algorithms for finite metric spaces, it is common
to assume that the distance between two points
can be computed in constant time, and complexity
bounds are expressed only in terms of the number
of points of the metric space.
We introduce a different model, where we assume
that the computation of a single distance is
an expensive operation and consequently, the goal is
to minimize the number of such distance queries.
This model is motivated by metric spaces
that appear in the context of topological data analysis.

We consider two standard operations on metric spaces,
namely the construction of a $(1+\epsilon)$-spanner
and the computation of an approximate nearest
neighbor for a given query point.
In both cases, we partially explore the metric space
through distance queries and infer lower and upper bounds
for yet unexplored distances through triangle inequality.
For spanners, we evaluate several exploration strategies
through extensive experimental evaluation.
For approximate nearest neighbors, we prove that our
strategy returns an approximate nearest neighbor
after a logarithmic number of distance queries.
\end{abstract}

\begin{document}

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\ccsdesc[500]{Theory of computation~Random search heuristics}

\copyrightyear{2019}
\acmYear{2020}
\acmConference[SOSA '20]{SIAM Symposium on Simplicity in Algorithms}{January 06--07, 2020}{Salt Lake City, UT, USA}
\acmBooktitle{SIAM Symposium on Simplicity in Algorithms (SOSA '20), January 06--07, 2020, Salt Lake City, UT, USA}
%\acmPrice{15.00}
%\acmDOI{10.1145/3295500.3356188}
%\acmISBN{978-1-4503-6229-0/19/11}

\maketitle

@@

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even a constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric \cite{rubner2000earth}.

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see Figure \ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

@@

\myparagraph{Spanners and known constructions}

Let $(\metricspace, \dist)$ be a finite metric space with $n$ points. 
One way to encode the metric space is a complete weighted graph on $\metricspace$,
where the weights correspond to the distances of the points.
A subgraph $G$ of this graph is called a \emph{$(1+\eps)$-spanner} for $(\metricspace,\dist)$ 
if for any pair of points $(u,v)$,
the shortest path distance $d_{uv}$ of $u$ and $v$ in $G$ satisfies $d_{u,v}\leq (1+\eps)\dist(u,v)$.
In other words, the shortest path metric of $G$ is a good approximation of the actual distance for every pair of points.
Clearly, it is a necessary condition that $G$ is connected, hence every spanner must have at least $n-1$ edges.

@@

\myparagraph{Well-Separated Pair Decomposition}
An alternative algorithm for computing a spanner is based  on well-separated pair decomposition (WSPD).
Let us recall the definition.  Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

On the other hand, spanner constructions based on WSPD~\cite{cal-kos-wspd,hm-fast} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a \espanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

@@

We performed two types of experiments: on points sampled from a low-dimensional Euclidean
space and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See Section \ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

Figure \ref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners produces
spanners of the same quality as the standard greedy algorithm,
\bgrdy and all variants of \brndm
perform significantly better than both variants of \textsc{BlindQuasiSorted}.

\ignore{
As for different variants of the \brndm algorithm,
we note that their performance is almost the same, and the algorithm works
significantly better than \textsc{QuasiSorted} variants, but obviously
worse than the blind greedy variant. There is a consistent, though small,
difference between the variants that do not force lower bounds first and
the other two variants of the \brndm.  (see Figure \ref{fig:blind_rbr_variants}). 
}

@@

From our experiments, we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners. More data from experiments in $\RR^n$ can be found in
Section \ref{sec:spanners_euclidean_addendum}.

@@

The other data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape data set; we only used
persistence in dimension 0, since the diagrams in this dimension are larger in size.
For the \dtype{perturbed} data set,
we added random noise to points of the original shapes,
creating 10 modified versions of each of them;
then we computed persistence diagrams in dimension 0 of these modified
shapes.
Persistence diagrams were computed with the \libraryname{DIPHA}
library \cite{bauer2014distributed}. 
Wasserstein distance were computed approximately
with the software library \hera using a relative error of $0.01$.
We restricted to approximate computation because computing the
Wasserstein distance exactly (using the Hungarian algorithm)
has been reported to be very expensice~\cite{kerber2017geometry}.

Sparseness values of \bgrdy spanner on the whole \dtype{original} data set
for different values of $\eps$ and $q$ are presented in Table \ref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
the advantage of our algorithm fades away. A plausible explanation
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in size of the greedy spanner.
For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
computations for $\eps = 0.1$; for a $1.5$-approximation, we
reduce the number of distances computed by a factor of 10.

@@

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in Table \ref{tbl:mcgill_original_blind_greedy_spanner_timing}.
We compared our approach for a fixed
$\eps$ with the time to compute all pairwise-distance between the
persistence diagrams using \hera, using the same approximation parameter
$\eps$. Let us emphasize that the diagrams we have are rather small,
with average cardinality
of 350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time,
and we expect this advantage to grow for diagrams of larger cardinality.

In order to see the growth of the spanner size, we use the \dtype{perturbed}
data set, drawing from it random samples.
We plot the sparseness of the spanners obtained by \grdy
and \bgrdy algorithms in Figure \ref{fig:spanner_sparseness_mcgill_perturbed}
for different values of $\eps$, always using the $\wsdist{1}$ distance.
As expected, the ratio of computed distances decreases as the number of diagrams
grows. In Figure \ref{fig:spanner_ratio_mcgill_perturbed} we show
the growth of the spanner size relative to the number of diagrams.
For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
linear growth) for larger values of $n$.

@@

The full proof is given in the next section,
we summarize the main ideas:
first, we notice that points ``far away'' from the current candidate $c$ 
can be disregarded in the algorithm. More precisely, if $p_i$
is outside of the ball of radius $2\dist(c, q)$ centered at $c$,
the lower bound for $p_i$ will be so large that
the distance to of $p_i$ to $q$ does not have to be queried.
Second, whenever a distance of $q$ to some $p_j$ is computed,
we can disregard all points ``very close'' to $p_j$.
Precisely, points in a ball of radius $\eps\dist(p_j,q)$
will not invalidate the current candidate 
to be an approximate nearest neighbor and hence,
no distance query for such points is needed.

@@

The proof strategy fails for $\eps=0$ because the $\eps$-ball
around $p_j$ as described above degenerates, and the packing argument fails.
Indeed, as the example in Figure \ref{fig:exact_bad_example} (in Section~\ref{sec:proof_ann}) 
shows, there are point sets where the expected number
of distance computations for exact nearest neighbour is linear.

@@

\section{Proof of Theorem \ref{thm:ann_bound}}
\label{sec:proof_ann}

Lemma \ref{lem:bound_lemma}
can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
further distance computations of points very close to $p_i$ or very far from $p_i$
will be avoided.

\definecolor{uququq}{rgb}{0.25,0.25,0.25}
\definecolor{xdxdff}{rgb}{0.49,0.49,1}
\definecolor{qqqqff}{rgb}{0,0,1}
\begin{figure}[ht!]
    \centering
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.5cm,y=0.5cm]
%\draw[->,color=black] (-1.5,0) -- (17,0);
%\foreach \x in {,2,4,6,8,10,12,14,16}
%\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
%\draw[->,color=black] (0,-7.5) -- (0,9.2);
%\foreach \y in {-6,-4,-2,2,4,6,8}
%\draw[shift={(0,\y)},color=black] (2pt,0pt) -- (-2pt,0pt) node[left] {\footnotesize $\y$};
%\draw[color=black] (0pt,-10pt) node[right] {\footnotesize $0$};
\clip(-1.5,-7.5) rectangle (17,9.2);
\draw [fill=black,fill opacity=0.5] (8.52,0.98) circle (1.02cm);
\draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
\draw [fill=black,fill opacity=0.95] (6.12,3.39) circle (0.42cm);
\draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
\draw [line width=2pt] (3.51,0.63) circle (1.9cm);
\draw [dash pattern=on 2pt off 2pt] (3.51,0.63) circle (1.49cm);
\draw (3.51,0.63)-- (0.62,1.33);
\draw (3.51,0.63)-- (6.12,3.39);
\draw (8.52,0.98)-- (3.51,0.63);
\draw (8.52,0.98)-- (14.89,-3.85);
\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps)$};
\begin{scriptsize}
\fill [color=qqqqff] (3.51,0.63) circle (1.5pt);
\draw[color=qqqqff] (3.71,0.94) node {$q$};
\fill [color=xdxdff] (0.62,1.33) circle (1.5pt);
\draw[color=black] (3.24,1.41) node {$r_1/(1+\eps)$};
\fill [color=qqqqff] (8.52,0.98) circle (1.5pt);
\draw[color=qqqqff] (8.9,1.29) node {$p_2$};
\fill [color=uququq] (6.48,0.83) circle (1.5pt);
\fill [color=qqqqff] (7.83,2.21) circle (1.5pt);
\draw[color=qqqqff] (8.21,2.52) node {$p_3$};
\fill [color=qqqqff] (7.65,-3.14) circle (1.5pt);
\draw[color=qqqqff] (8.01,-2.82) node {$p_4$};
\fill [color=qqqqff] (6.62,0.6) circle (1.5pt);
\draw[color=qqqqff] (6.98,0.92) node {$p_5$};
\fill [color=qqqqff] (1.49,-2.39) circle (1.5pt);
\draw[color=qqqqff] (1.87,-2.08) node {$p_6$};
\fill [color=qqqqff] (1.95,-2.65) circle (1.5pt);
\draw[color=qqqqff] (2.31,-2.33) node {$p_7$};
\fill [color=qqqqff] (2.3,-2.75) circle (1.5pt);
\draw[color=qqqqff] (2.68,-2.43) node {$p_8$};
\fill [color=qqqqff] (2.79,-2.9) circle (1.5pt);
\draw[color=qqqqff] (3.17,-2.57) node {$p_9$};
\fill [color=xdxdff] (14.89,-3.85) circle (1.5pt);
\fill [color=uququq] (5.84,3.09) circle (1.5pt);
\fill [color=xdxdff] (4.52,4.29) circle (1.5pt);
\draw[color=xdxdff] (4.89,4.61) node {$p_1$};
\end{scriptsize}
\end{tikzpicture}
\caption{First two steps of the ANN algorithm. 
%    First $p_1$ is 
%chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
%distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered 
%at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
%is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$
%in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. 
%This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the
%brute-force algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis. 
The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
\label{fig:ann_illustration}
\end{figure}

In Figure \ref{fig:ann_illustration} we illustrate this. First, $p_1$ is
chosen as the current candidate, and we must compute $\dist(p_2, q)$.
After that the algorithm will not compute
distance to any of the points inside the heavily shaded ball or outside
the lightly shaded ball that are centered at $p_2$,
because their lower bounds allow us to discard them.

\begin{lemma}
\label{lem:sequence_lemma}
Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
\[\left(\frac{4(2+\eps) v_k}{\eps v_\ell}\right)^{d}\]
distances to $q$.
\end{lemma}
\begin{proof}
By the first part of Lemma \ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
is computed, the second part of Lemma \ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma \ref{lem:packing_lemma}.
See Figure \ref{fig:ann_illustration}.
\end{proof}

A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
the number of computed distances is a constant (since $v_k=v_\ell$). 
This means that to prove Theorem \ref{thm:ann_bound}, it would suffice to show
that the candidate changes only a logarithmic number of times in expectation.
Let us consider Figure \ref{fig:ann_illustration} again. Notice that the point $p_5$, which
is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
$p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are farther from 
$q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the 
candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis,
and we need to modify it slightly.

\begin{proof} (of Theorem \ref{thm:ann_bound})
%
In the sequence $p_1,\ldots,p_n$, let $p_k$ be a point 
such that $r_i<r_k$ for all $1\leq i\leq k-1$. We call an element
of this form a \emph{minimum} of the sequence. 
A standard backwards analysis argument~\cite{seidel-backwards} shows that the probability
of $p_k$ being a minimum is at most $1/k$, so that the number of minima
in the sequence is $O(\log n)$ in expectation.
